---
title: 
output:
  html_document:
    toc: true
    toc_float: true
---

### Introduction to R
We will be using real oceanographic datasets to learn various skills, such as manipulating and merging datasets, plotting profiles, making maps, and anything else that might be useful.   

Before we start with that, let's take a few minutes to start with the basics for people that have never used R before. First off, let's introduce you to entering data into R.    

## Comments 
You will soon notice how some of the code that we have looked at has '#' in front of it. By starting a line with this symbol, we are telling R to not run this line as part of the code. This allows us to leave notes for ourselves (comments) that can be incredibly helpful in keeping track of what we are doing, especially if we haven't used the script in a long time. It is in your best interests to make this a habit...you will appreciate it later!

### Values and Vectors
You can store values, be they numbers or letters, and assign names to them.   
As an example, let's create a series of names.  
We will do this using the function **c()**, which combines values into a vector and save it as "x".  
These data are in the class 'character'.  
Anything inside quotation marks "" is a string.  
```{r}
# create vector
x <- c("Craig", "Mark", "Debora", "Alyson", "Uta")
# print out vector
x
# check vector class
class(x)
```
Next, let's create a vector of numbers, also called x.  
Unlike the previous example, numbers aren't inside quotation marks.  
This makes each of them a 'numeric', which can be used in mathmatical expressions.
```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
x

# check vector class
class(x)

# add up all the numbers
sum(x)
```
<br>  
What if you put the numbers in quotation marks?
```{r}
x <- c('1','2','3','4','5','6','7','8','9','10')
x

# check vector class
class(x)

```
It didn't work, did it?  

This happened because we tried to do math on values that were stored as characters, which R will not do.  
Whenever you are working with data, it is always good to check the class of the values to make sure they are in the correct format.   
<br>

If you are like me, you don't like entering in a lot of values individually such as creating a vector of depths.
Here are a few tricks to make it easier
```{r}
# recreate a set of identical values
x <- c(rep(10,10))
x

# create a sequence of values from 1 to 10 by 1
x <- c(seq(1,10,1))
x

# create a sequence of values from 1 to 50 by 5
x <- c(seq(0,51,5))
x
```

We can also get fancy by combining multiple sequences to create a complex vector
```{r}  
# create vector
x <- c(seq(0,1,0.1), seq(2,10,1), seq(20,100,10))
x
```
<br>  

### Dataframes 
If you are working with an observational dataset, such as cruise data, it will likely be in the form of a dataframe.  
A dataframe is a collection of vectors identified by the a name. This is very similar to what you are likely used to works with excel spreadsheets.

Let's start off by creating our dataset from a PnB cruise simulating a 100 m cast comprised of depth (0-100m), and two hypothetical parameters, paramA and paramB:

```{r}
# create a vector for the name of the cruise called "PnB"
cruise <- c(rep("PnB",101))
# create vector of depth
depth <- c(seq(0,100))
# create 101 randomn values of 'paramA' from values ranging between 0 and 10 
paramA <- c(runif(101, min=0, max=10)) 
# create 101 randomn values of 'paramB' from values ranging between 1 and 100 
paramB <- c(sample(1:100,101, replace=TRUE)) 
# combine the vectors into a dataframe and do not turn strings into factors
df <- data.frame(cruise, depth, paramA, paramB, stringsAsFactors=FALSE) 
```
<br>   
We can now check the structure of our dataset by using the **str()** function which will tell us the class of each of our vectors
```{r}
str(df) # check structure of dataframe
```

### Saving a dataframe
Suppose we want to save our dataframe so that we can either send it to a collaborator or open it in another program.  

The easiest way is to save it as a .csv file.  
First, we need to figure out where it is going to be saved. The best way to do this is to check our working directory using the function **getwd()**  
```{r}
# get the working directory
getwd()
```
If we want it to go somewhere else, we can either change the working directory using the function **setwd()**  
<br>  
Then we use the function **write.csv()** to create the file that will be saved in that directory
```{r}
# write a csv file
write.csv(df, file = "MyFirstDataframe.csv", row.names=F) 
```
If we do not want to chnage our working directory an alternative approach is to specify the location of the file in its name such as  
'file= "~/Users/paulmatson/tmp/MyFirstDataframe.csv"'  

<br>  

### Loading data
One of the easiest ways to load in a table of data is to use the function **read.table()**  

Read in your file called "MyFirstDataframe.csv" and store it as 'data'.  
If the file is in your working directory, you only need to specify the name, otherwise you will need to provide the directory path to find it.
```{r}
# read in file and store as 'data'
# the 'header' argument specifies where your file contains a header row
# the 'sep' argument specifies how your values are separated; for .csv files a comma is used.
data <- read.table("MyFirstDataframe.csv", header=TRUE, sep=",")
```
Now you should have the same dataframe stored as both 'df' and 'data'.



